title: "City Comparison Dashboard – Income & Rent (2021)"
author: "Max Desroches"
date: "r format(Sys.Date(), '%Y-%m-%d')"
output:
html_document:
toc: true
toc_float: true
number_sections: false
theme: flatly
highlight: tango
df_print: paged
code_folding: show


```{r setup, include=FALSE}
knitr::opts_chunk$set(
echo = TRUE,
message = FALSE,
warning = FALSE,
fig.width = 10,
fig.height = 5.5,
fig.retina = 2
)

library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(forcats)
library(stringr)
library(here)
library(scales)

# ---- Project scripts ----

source(here("scripts", "utils.R"))
source(project_path("scripts", "load_data.R"))

if (file.exists(project_path("scripts", "cleaning.R")))  source(project_path("scripts", "cleaning.R"))
if (file.exists(project_path("scripts", "transform.R"))) source(project_path("scripts", "transform.R"))


# ---- Formatting helpers (CAD) ----

cad0 <- label_dollar(prefix = "$", big.mark = ",", accuracy = 1) # $1,234
cadK <- label_dollar(prefix = "$", big.mark = ",", accuracy = 1, scale = 1e-3, suffix = "K") # $65K
pct1 <- label_percent(accuracy = 0.1)

# ---- Dashboard theme ----

dashboard_theme <- function() {
theme_minimal(base_size = 12) +
theme(
plot.title    = element_text(face = "bold", size = 16, margin = margin(b = 6)),
plot.subtitle = element_text(size = 11, color = "gray30", margin = margin(b = 10)),
plot.caption  = element_text(size = 9, color = "gray50", margin = margin(t = 10)),
axis.title    = element_text(face = "bold"),
axis.text     = element_text(color = "gray20"),
panel.grid.minor   = element_blank(),
panel.grid.major.x = element_blank(),
legend.position    = "bottom"
)
}

# ---- KPI cards CSS ----

card_css <- "

<style>
  .cards { display:flex; gap:14px; flex-wrap:wrap; margin: 10px 0 16px 0; }
  .card  { background:#ffffff; border:1px solid #e9ecef; border-radius:14px;
           padding:14px 16px; min-width:220px; box-shadow: 0 6px 18px rgba(0,0,0,0.04); }
  .kpi   { font-size:22px; font-weight:700; margin: 0; }
  .lbl   { font-size:12px; color:#6c757d; margin: 0; }
  .note  { font-size:12px; color:#6c757d; margin-top:6px; }
</style>

"

# ---- visuals folder + helper ----

dir.create(project_path("visuals"), showWarnings = FALSE, recursive = TRUE)
short_city <- function(x) stringr::str_trunc(x, width = 22, side = "right")

```

# Overview

This dashboard summarizes income vs rent patterns across cities and creates an affordability ranking using a rent-to-income ratio.

# Data loading
```{r}
income_raw <- load_income_data("income_2021.csv")
rent_raw   <- load_rent_data("rent_2021.csv")

glimpse(income_raw)
glimpse(rent_raw)

cat("Rows in income:", nrow(income_raw), "\n")
cat("Rows in rent  :", nrow(rent_raw), "\n")

```

# Data prep
```{r}
income_clean <- clean_income_data(income_raw)
rent_clean   <- clean_rent_data(rent_raw)

city_df <- combine_income_rent(income_clean, rent_clean)

# Affordability metric

city_df <- city_df |>
mutate(rent_to_income_ratio = (avg_monthly_rent_2021 * 12) / median_income_2021)

# Rankings (full list)

city_rank <- city_df |>
filter(!is.na(rent_to_income_ratio)) |>
arrange(rent_to_income_ratio) |>
mutate(affordability_rank = row_number())

# Ordered views (useful when dataset has few cities)

city_rank_asc  <- city_rank |> arrange(rent_to_income_ratio)
city_rank_desc <- city_rank |> arrange(desc(rent_to_income_ratio))

# Correlation

cor_income_rent <- cor(
city_df$median_income_2021,
city_df$avg_monthly_rent_2021,
use = "complete.obs"
)

# KPIs

n_cities   <- nrow(city_rank)
med_income <- median(city_rank$median_income_2021, na.rm = TRUE)
med_rent   <- median(city_rank$avg_monthly_rent_2021, na.rm = TRUE)
med_ratio  <- median(city_rank$rent_to_income_ratio, na.rm = TRUE)

```

```{r}
cat(card_css)

cat('<div class="cards">')
cat(sprintf(
'<div class="card"><p class="lbl">Cities analyzed</p><p class="kpi">%s</p><p class="note">After cleaning & matching</p></div>',
comma(n_cities)
))
cat(sprintf(
'<div class="card"><p class="lbl">Median income (2021)</p><p class="kpi">%s</p><p class="note">Assumed CAD</p></div>',
dollar(med_income, prefix="$", big.mark=",", accuracy=1)
))
cat(sprintf(
'<div class="card"><p class="lbl">Median rent / month (2021)</p><p class="kpi">%s</p><p class="note">Assumed CAD</p></div>',
dollar(med_rent, prefix="$", big.mark=",", accuracy=1)
))
cat(sprintf(
'<div class="card"><p class="lbl">Median rent-to-income</p><p class="kpi">%s</p><p class="note">Lower is better</p></div>',
percent(med_ratio, accuracy=0.1)
))
cat('</div>')

```

# Distributions
## Income distribution
```{r}
ggplot(city_rank, aes(x = median_income_2021)) +
geom_histogram(bins = 30, fill = "#60a5fa", alpha = 0.85) +
scale_x_continuous(labels = cadK) +
labs(
title = "Income distribution",
subtitle = "Median city income (2021) — shown in CAD, scaled to thousands",
x = "Median income (CAD, thousands)",
y = "Number of cities",
caption = "City-Comparison-Canada project data"
) +
dashboard_theme()
```

## Rent distribution
```{r}
ggplot(city_rank, aes(x = avg_monthly_rent_2021)) +
geom_histogram(bins = 30, fill = "#93c5fd", alpha = 0.85) +
scale_x_continuous(labels = cad0) +
labs(
title = "Rent distribution",
subtitle = "Average monthly rent (2021) — CAD",
x = "Average monthly rent (CAD)",
y = "Number of cities",
caption = "City-Comparison-Canada project data"
) +
dashboard_theme()
```

# Relationship: income vs rent
```{r}
ggplot(city_rank, aes(x = median_income_2021, y = avg_monthly_rent_2021)) +
geom_point(color = "#1f77b4", alpha = 0.65, size = 2) +
geom_smooth(method = "lm", se = TRUE, linetype = "dashed",
color = "#1f77b4", fill = "#1f77b4", alpha = 0.15) +
scale_x_continuous(labels = cadK) +
scale_y_continuous(labels = cad0) +
labs(
title = "Income vs rent (by city)",
subtitle = paste0("Trend line (linear model). Correlation: ", round(cor_income_rent, 3)),
x = "Median income (CAD, thousands)",
y = "Average monthly rent (CAD)",
caption = "Correlation computed using complete observations"
) +
dashboard_theme()

```

# Most affordable cities (2021)
```{r}
city_rank_asc |>
mutate(city_lbl = short_city(city),
city_lbl = forcats::fct_reorder(city_lbl, rent_to_income_ratio)) |>
ggplot(aes(x = city_lbl, y = rent_to_income_ratio)) +
geom_col(fill = "#2ca02c", alpha = 0.85) +
geom_text(aes(label = percent(rent_to_income_ratio, accuracy = 0.1)),
hjust = -0.05, size = 3.3) +
coord_flip(clip = "off") +
scale_y_continuous(labels = pct1, expand = expansion(mult = c(0, 0.18))) +
labs(
title = "Most affordable cities (2021)",
subtitle = "Full ranking (lower rent-to-income ratio is better)",
x = NULL,
y = "Rent-to-income ratio",
caption = "Metric: (avg_monthly_rent_2021 * 12) / median_income_2021"
) +
dashboard_theme()

```

# Least affordable cities (2021)
```{r}
city_rank_desc |>
mutate(city_lbl = short_city(city),
city_lbl = forcats::fct_reorder(city_lbl, rent_to_income_ratio)) |>
ggplot(aes(x = city_lbl, y = rent_to_income_ratio)) +
geom_col(fill = "#2563eb", alpha = 0.85) +
geom_text(aes(label = percent(rent_to_income_ratio, accuracy = 0.1)),
hjust = -0.05, size = 3.3) +
coord_flip(clip = "off") +
scale_y_continuous(labels = pct1, expand = expansion(mult = c(0, 0.18))) +
labs(
title = "Least affordable cities (2021)",
subtitle = "Full ranking (higher rent-to-income ratio is worse)",
x = NULL,
y = "Rent-to-income ratio",
caption = "Metric: (avg_monthly_rent_2021 * 12) / median_income_2021"
) +
dashboard_theme()

```

# Tables
```{r}
city_rank |>
select(affordability_rank, city, median_income_2021, avg_monthly_rent_2021, rent_to_income_ratio) |>
mutate(
median_income_2021 = dollar(median_income_2021, prefix="$", big.mark=",", accuracy=1),
avg_monthly_rent_2021 = dollar(avg_monthly_rent_2021, prefix="$", big.mark=",", accuracy=1),
rent_to_income_ratio = percent(rent_to_income_ratio, accuracy=0.1)
)

```

# Saving outputs (PNG)
```{r}
# Save cleaned dataset

output_path <- project_path("data", "city_income_rent_clean.csv")
readr::write_csv(city_rank, output_path)

# Create visuals folder (IMPORTANT)
dir.create(project_path("visuals"), showWarnings = FALSE, recursive = TRUE)

# Save key plots to visuals/

p_income_rent <- ggplot(city_rank, aes(x = median_income_2021, y = avg_monthly_rent_2021)) +
geom_point(color = "#1f77b4", alpha = 0.65, size = 2) +
geom_smooth(method = "lm", se = TRUE, linetype = "dashed",
color = "#1f77b4", fill = "#1f77b4", alpha = 0.15) +
scale_x_continuous(labels = cadK) +
scale_y_continuous(labels = cad0) +
labs(
title = "Income vs Rent (2021)",
subtitle = paste0("Cities: ", nrow(city_rank), " | Correlation: ", round(cor_income_rent, 3)),
x = "Median income (CAD, thousands)",
y = "Average monthly rent (CAD)",
caption = "Metric uses complete observations only"
) +
dashboard_theme()

ggsave(
filename = project_path("visuals", "01_income_vs_rent.png"),
plot = p_income_rent,
width = 12, height = 6.5, dpi = 320, bg = "white"
)

p_affordable <- city_rank_asc |>
mutate(city_lbl = short_city(city),
city_lbl = forcats::fct_reorder(city_lbl, rent_to_income_ratio)) |>
ggplot(aes(x = city_lbl, y = rent_to_income_ratio)) +
geom_col(fill = "#2ca02c", alpha = 0.85) +
geom_text(aes(label = percent(rent_to_income_ratio, accuracy = 0.1)),
hjust = -0.05, size = 3.3) +
coord_flip(clip = "off") +
scale_y_continuous(labels = pct1, expand = expansion(mult = c(0, 0.18))) +
labs(
title = "Most affordable cities (2021)",
subtitle = "Full ranking (lower rent-to-income ratio is better)",
x = NULL, y = "Rent-to-income ratio"
) +
dashboard_theme()

ggsave(
filename = project_path("visuals", "02_affordability_most_affordable.png"),
plot = p_affordable,
width = 12, height = 7, dpi = 320, bg = "white"
)

p_unaffordable <- city_rank_desc |>
mutate(city_lbl = short_city(city),
city_lbl = forcats::fct_reorder(city_lbl, rent_to_income_ratio)) |>
ggplot(aes(x = city_lbl, y = rent_to_income_ratio)) +
geom_col(fill = "#2563eb", alpha = 0.85) +
geom_text(aes(label = percent(rent_to_income_ratio, accuracy = 0.1)),
hjust = -0.05, size = 3.3) +
coord_flip(clip = "off") +
scale_y_continuous(labels = pct1, expand = expansion(mult = c(0, 0.18))) +
labs(
title = "Least affordable cities (2021)",
subtitle = "Full ranking (higher rent-to-income ratio is worse)",
x = NULL, y = "Rent-to-income ratio"
) +
dashboard_theme()

ggsave(
filename = project_path("visuals", "03_affordability_least_affordable.png"),
plot = p_unaffordable,
width = 12, height = 7, dpi = 320, bg = "white"
)

```

# Conclusion
This analysis examined the relationship between **city-level income and rent** across selected Canadian cities using 2021 data.

The results show a **strong positive relationship** between income and rent, indicating that cities with higher median incomes also tend to have higher housing costs. However, income growth does not fully offset rent increases in all locations, leading to meaningful differences in **housing affordability**.

Using the **rent-to-income ratio** as an affordability metric highlights clear contrasts between cities. Mid-sized cities such as **Quebec City, Calgary, and Edmonton** appear more affordable, while larger metropolitan areas like **Toronto and Vancouver** rank among the least affordable.

Overall, this analysis demonstrates how simple, transparent metrics can provide valuable insights into cost-of-living differences across cities. The framework can be extended in future work by incorporating additional variables such as population growth, labor market indicators, or multi-year trends to support deeper policy or investment analysis.




